### 2018-03-04 doc/ms.tex
A document

\documentclass[a4paper, 12pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage[left=2cm, right=2cm, bottom=3cm, top=2cm]{geometry}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{natbib}
\usepackage{transparent}

\title{Information Loss from the Whittle Likelihood}
\author{Brendon J. Brewer}
\date{2018}

\begin{document}
\maketitle

\abstract{\noindent I investigate the information content of a time series
          dataset for inferring a parameter of interest. }

% Need this after the abstract
\setlength{\parindent}{0pt}
\setlength{\parskip}{1em}

\section{A Damped, Stochastically-Excited Oscillator}

\begin{thebibliography}{999}
\bibitem[Foreman-Mackey et al.(2017)]{celerite}
Foreman-Mackey, D., Agol, E., Ambikasaran, S. and Angus, R., 2017. Fast and scalable Gaussian process modeling with applications to astronomical time series. The Astronomical Journal, 154(6), p.220.
\end{thebibliography}

\end{document}

### 2018-03-04 src/WhiteNoise.cpp
Removed stray comment

@@ -98,7 +98,7 @@ void WhiteNoise::print(std::ostream& out) const
     for(size_t i=0; i<N; ++i)
         out << y[i] << ' ';
 }
-9.60 +- 0.15
+
 double WhiteNoise::parameter_distance(const WhiteNoise& s1, const WhiteNoise& s2)
 {
     return std::abs(log(s2.sigma) - log(s1.sigma));

### 2018-03-05 doc/ms.tex
Results
Another table
Extra space between columns
More nice tables

@@
 \usepackage{amssymb}
+\usepackage{booktabs} % For nicer tables
 \usepackage{color}
@@
 \usepackage{transparent}
+
+\renewcommand{\arraystretch}{1.2} % More space between table rows
 
@@
 
+\begin{table}[!ht]
+\centering
+\begin{tabular}{@{}l@{\hspace{3em}}l@{\hspace{2em}}l@{}}
+\toprule
+$N$         &       Likelihood      &  Mut. Inf. (nats) \\
+\hline
+100         &       Exact           & $2.605 \pm 0.053$ \\
+100         &       Whittle         & $2.052 \pm 0.055$ \\
+\bottomrule
+\end{tabular}
+\caption{\label{tab:oscillator_results}}
+\end{table}
+
+\begin{table}[!ht]
+\centering
+\begin{tabular}{@{}l@{\hspace{3em}}l@{}}
+\toprule
+$N$         &       $\Delta$ Mut. Inf. (nats) \\
+\hline
+100         &       $0.553 \pm 0.056$ \\
+\bottomrule
+\end{tabular}
+\caption{The difference in mutual information between the exact and
+Whittle likelihoods. This quantifies the size of the advantage of the exact
+likelihood.
+The error bars are smaller than what you might expect from
+taking differences in Table~\ref{tab:oscillator_results}, due to
+the common stream of datasets.\label{tab:oscillator_differences}}
+\end{table}
+
 \begin{thebibliography}{999}

### 2018-03-05 doc/ms.tex
Wrote down the ODE

@@
 \section{A Damped, Stochastically-Excited Oscillator}
+Consider the following ordinary differential equation, which describes
+a damped and excited oscillator:
+\begin{align}
+\frac{d^2y}{dt^2} + \frac{\omega_0}{Q}\frac{dy}{dt}
+        + \omega_0^2\frac{dy}{dt} &= \epsilon(t).\label{eqn:ode}
+\end{align}
+The function $y(t)$ is the resulting signal, and the parameters $\omega_0$
+are the angular frequency (that the signal would have if
+the driving force $\epsilon(t)$ were zero) and $Q$ is the quality factor.
+If the driving force $\epsilon{t}$ is only defined probabilistically (i.e.,
+it is specified to be white noise) then Equation~\ref{eqn:ode} becomes a
+stochastic differential equation whose solution is a probability distribution
+for $y(t)$.
 

### 2018-03-05 doc/ms.tex
More results
New title

@@
 
-\title{Information Loss from the Whittle Likelihood}
+\title{Bayesian inference from time series: Information loss and the Whittle likelihood}
 \author{Brendon J. Brewer}
@@
 \hline
+50          &       Exact           & \\
+50          &       Whittle         & $1.957 \pm 0.055$ \\
 100         &       Exact           & $2.605 \pm 0.053$ \\
 100         &       Whittle         & $2.052 \pm 0.055$ \\
+200         &       Exact           &   \\
+200         &       Whittle         &   \\
 \bottomrule
@@
 \hline
+50          & \\
 100         &       $0.553 \pm 0.056$ \\
+200         & \\
 \bottomrule

### 2018-03-05 doc/ms.tex
Fixed typo

@@
 the driving force $\epsilon(t)$ were zero) and $Q$ is the quality factor.
-If the driving force $\epsilon{t}$ is only defined probabilistically (i.e.,
+If the driving force $\epsilon(t)$ is only defined probabilistically (i.e.,
 it is specified to be white noise) then Equation~\ref{eqn:ode} becomes a

### 2018-03-05 doc/ms.tex
Another result

@@
 \hline
+20          &       Exact           & \\
+20          &       Whittle         & $1.491 \pm 0.055$\\ 
 50          &       Exact           & \\

### 2018-03-05 doc/ms.tex
Wrote down C(tau)

--- 2018-03-05 doc/ms.tex
@@
 
+We assume that the probability distribution for $\epsilon(t)$ is indeed white
+noise. In this case, the probability distribution for $y(t)$ given the
+parameters is a gaussian process with mean function of zero and
+covariance function given by
+\begin{align}
+C(\tau) &= A^2\left[\cos(\eta\omega_0\tau)
+                + \frac{1}{2\eta Q}\sin(\eta\omega_0\tau)\right]
+\end{align}
+where $\eta = \sqrt{|1 - 1/(4Q^2)|}$, $\tau$ is a time separation,
+and $Q > 1/2$ (i.e., we assume the oscillation is not overdamped).
+An equivalent expression is given by \citet{anderson} and
+\citet{celerite}.
+
 \begin{table}[!ht]
@@
 \begin{thebibliography}{999}
+\bibitem[Anderson et al.(1990)]{anderson}
+Anderson, Edwin R., Thomas L. Duvall Jr, and Stuart M. Jefferies. "Modeling of solar oscillation power spectra." The Astrophysical Journal 364 (1990): 699-705.
+
 \bibitem[Foreman-Mackey et al.(2017)]{celerite}

### 2018-03-05 doc/ms.tex
More writing

@@
 
-\title{Bayesian inference from time series: Information loss and the Whittle likelihood}
+\title{Bayesian Inference from Time Series: Information Loss and the Whittle Likelihood}
 \author{Brendon J. Brewer}
@@
 \end{align}
-where $\eta = \sqrt{|1 - 1/(4Q^2)|}$, $\tau$ is a time separation,
+where $\eta = \sqrt{1 - 1/(4Q^2)}$, $\tau$ is a time separation,
 and $Q > 1/2$ (i.e., we assume the oscillation is not overdamped).
-An equivalent expression is given by \citet{anderson} and
-\citet{celerite}.
+A more general expression which allows for $Q \leq 1/2$
+is given by \citet{anderson} and \citet{celerite}.
+In the limit of large $Q$, the covariance function is approximately
+the product of an oscillation and an exponential decay:
+\begin{align}
+C(\tau) &= A\exp\left(-\frac{\omega_0\tau}{2Q}\right)\cos(\omega_0\tau)
+\end{align}
+which was the form assumed by \citep{brewer_stello}.
 
@@
 \bibitem[Anderson et al.(1990)]{anderson}
-Anderson, Edwin R., Thomas L. Duvall Jr, and Stuart M. Jefferies. "Modeling of solar oscillation power spectra." The Astrophysical Journal 364 (1990): 699-705.
+Anderson, Edwin R., Thomas L. Duvall Jr, and Stuart M. Jefferies. ``Modeling of solar oscillation power spectra.'' The Astrophysical Journal 364 (1990): 699-705.
+
+\bibitem[Brewer and Stello(2009)]{brewer_stello}
+Brewer, Brendon J., and Dennis Stello. ``Gaussian process modelling of asteroseismic data.'' Monthly Notices of the Royal Astronomical Society 395, no. 4 (2009): 2226-2233.
 

### 2018-03-05 doc/ms.tex
PSD
More writing

@@
 
+\section{Introduction}
+
+In a recent paper, I introduced a computational technique based on Nested
+Sampling for estimating the entropy of probability distributions which are
+usually only accessible through sampling, such as posterior distributions.
+
 \section{A Damped, Stochastically-Excited Oscillator}
@@
                 + \frac{1}{2\eta Q}\sin(\eta\omega_0\tau)\right]
+    \label{eqn:oscillation_covariance}
 \end{align}
@@
 \end{align}
-which was the form assumed by \citep{brewer_stello}.
+which was the form assumed by \citet{brewer_stello}.
+
+The power spectral density corresponding to
+Equation~\ref{eqn:oscillation_covariance} is
+\begin{align}
+S(\omega) &= \sqrt{\frac{2}{\pi}}
+             \frac{S_0\omega_0^4}
+                  {(\omega^2 - \omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
+\end{align}
+where $S_0 = A/(\omega_0Q)$.
+
+\subsection{Inferring the period}
+The parameter $\omega_0$ is the angular frequency, related to the period
+$T$ by
+\begin{align}
+T &= \frac{2\pi}{\omega_0}.
+\end{align}
 
@@
 \end{tabular}
-\caption{\label{tab:oscillator_results}}
+\caption{The mutual information between the data and the log-period, for
+different numbers of datapoints $N$ and for different likelihoods (exact vs.
+Whittle).
+This is equal to the prior expected reduction in the entropy
+in going from the prior to the posterior.\label{tab:oscillator_results}}
 \end{table}
@@
 
+\bibitem[Brewer(2017)]{brewer}
+Brewer, Brendon J. ``Computing Entropies with Nested Sampling.'' Entropy 19, no. 8 (2017): 422.
+
 \bibitem[Brewer and Stello(2009)]{brewer_stello}
@@
 \bibitem[Foreman-Mackey et al.(2017)]{celerite}
-Foreman-Mackey, D., Agol, E., Ambikasaran, S. and Angus, R., 2017. Fast and scalable Gaussian process modeling with applications to astronomical time series. The Astronomical Journal, 154(6), p.220.
+Foreman-Mackey, Daniel, Eric Agol, Sivaram Ambikasaran, and Ruth Angus.
+``Fast and scalable Gaussian process modeling with applications to astronomical time series.'' The Astronomical Journal 154, no.
+6 (2017): 220.
 \end{thebibliography}

### 2018-03-05 doc/ms.tex
Writing about the oscillation data

@@
 which was the form assumed by \citet{brewer_stello}.
-
 The power spectral density corresponding to
@@
 \end{align}
+In this section we assume that it is the log-period $\log_{10}(T)$ that is
+of interest, and that the observations consist of $N$ measurements
+$\boldsymbol{y} = \{y_0, y_1, ..., y_{N-1}\}$,
+at times $\boldsymbol{t} = \{0,1,2,...,N-1\}$, observed with noise:
+\begin{align}
+y_i &\sim \textnormal{Normal}(y(t_i), \sigma^2).
+\end{align}
+The noise level $\sigma$ was set to unity.
+
+The unknown parameters are the oscillation parameters $(A, T, Q)$, to which
+we assigned the following priors:
+\begin{align}
+\ln A &\sim \textnormal{Normal}(0, 1) \\
+\log_{10} T &\sim \textnormal{Uniform}(\log_{10}(N) - 1, \log_{10}(N)) \\
+\ln Q &\sim \textnormal{Uniform}(\ln 1, \ln 1000).
+\end{align}
+We proceeded to compute the conditional entropy
+$H(\log_{10}(\tau) | \boldsymbol{y})$ using the approach of \citet{brewer},
+and hence the mutual information $I(\log_{10}(\tau) ; \boldsymbol{y})$.
+The results are given in Table~\ref{tab:oscillation_results}.
 

### 2018-03-05 doc/ms.tex
More citations and intro writing

@@
 usually only accessible through sampling, such as posterior distributions.
+
+In astronomy, time series analysis of variable sources has often been performed
+in the frequency domain \citep[e.g.][]{lomb, hatzes, helio}. However,
+a number of researchers have applied existing time series models from
+statistics \citep{brockwell_davis} or custom-made models to infer parameters
+directly from the time series data, without any use of the frequency domain
+\citep[e.g.][]{gregory_loredo, gregory, kelly, brewer_stello, farr}.
+
+Methods that analyse the time series data directly are often less convenient than frequency-domain fits, and can have a much greater computational cost,
+despite recent computational advances which have improved the matter
+considerably \citep[e.g.][]{hodlr, celerite}. Yet the question remains ---
+if we opt for the convenient frequency-domain methods, how much relevant
+information are we throwing away? That is the focus of this paper.
 
@@
 This is equal to the prior expected reduction in the entropy
-in going from the prior to the posterior.\label{tab:oscillator_results}}
+in going from the prior to the posterior.\label{tab:oscillation_results}}
 \end{table}
@@
 The error bars are smaller than what you might expect from
-taking differences in Table~\ref{tab:oscillator_results}, due to
-the common stream of datasets.\label{tab:oscillator_differences}}
+taking differences in Table~\ref{tab:oscillation_results}, due to
+the common stream of datasets.\label{tab:oscillation_differences}}
 \end{table}
@@
 \begin{thebibliography}{999}
+
+\bibitem[Ambikasaran et al.(2016)]{hodlr}
+Ambikasaran, Sivaram, Daniel Foreman-Mackey, Leslie Greengard, David W. Hogg, and Michael O’Neil. ``Fast direct methods for Gaussian processes.''
+IEEE transactions on pattern analysis and machine intelligence 38, no. 2 (2016): 252-265.
+
 \bibitem[Anderson et al.(1990)]{anderson}
@@
 \bibitem[Brewer and Stello(2009)]{brewer_stello}
-Brewer, Brendon J., and Dennis Stello. ``Gaussian process modelling of asteroseismic data.'' Monthly Notices of the Royal Astronomical Society 395, no. 4 (2009): 2226-2233.
+Brewer, Brendon J., and Dennis Stello. ``Gaussian process modelling of asteroseismic data.'' Monthly Notices of the Royal Astronomical Society 395, no. 4 
+(2009): 2226-2233.
+
+\bibitem[Brockwell and Davis(2013)]{brockwell_davis}
+Brockwell, P.J. and Davis, R.A., 2013. Time series: theory and methods. Springer Science \& Business Media.
+
+\bibitem[Christensen-Dalsgaard(2002)]{helio}
+Christensen-Dalsgaard, Jørgen. ``Helioseismology.'' Reviews of Modern Physics 74, no. 4 (2002): 1073.
+
+\bibitem[Farr et al.(2018)]{farr}
+Farr, Will M., Benjamin JS Pope, Guy R.
+Davies, Thomas SH North, Timothy R.
+White, Jim W.
+Barrett, Andrea Miglio et al.
+``Aldebaran b's temperate past uncovered in planet search data.'' arXiv preprint arXiv:1802.09812 (2018).
 
@@
 6 (2017): 220.
+
+\bibitem[Gregory(2005)]{gregory}
+Gregory, Philip C.
+``A Bayesian analysis of extrasolar planet data for HD 208487.''
+In AIP Conference Proceedings, vol. 803, no. 1, pp. 139-145. AIP, 2005.
+
+\bibitem[Gregory and Loredo(1992)]{gregory_loredo}
+Gregory, P. C., and Thomas J. Loredo. ``A new method for the detection of a periodic signal of unknown shape and period.'' The Astrophysical Journal 398 (1992): 146-168.
+
+\bibitem[Hatzes et al.(1993)]{hatzes}
+Hatzes, A.P. and Cochran, W.D., 1993. Long-period radial velocity variations in three K giants. The Astrophysical Journal, 413, pp.339-348.
+
+\bibitem[Kelly(2009)]{kelly}
+Kelly, Brandon C., Jill Bechtold, and Aneta Siemiginowska. ``Are the variations in quasar optical flux driven by thermal fluctuations?.'' The Astrophysical Journal 698, no. 1 (2009): 895.
+
+\bibitem[Lomb(1976)]{lomb}
+Lomb, Nicholas R. ``Least-squares frequency analysis of unequally spaced data.'' Astrophysics and space science 39, no. 2 (1976): 447-462.
+
 \end{thebibliography}

### 2018-03-05 include/AR1.h
AR(1)

@@ -1,0 +1,50 @@
+#ifndef WhittledAway_AR1_h
+#define WhittledAway_AR1_h
+
+// Includes
+#include <armadillo>
+#include <Eigen/Dense>
+#include <ostream>
+#include <vector>
+#include "Model.h"
+#include "RNG.h"
+
+namespace WhittledAway
+{
+
+class AR1 : public Model
+{
+    private:
+
+        // Mean level, beta, L
+        double mu, beta, L;
+
+        // Calculate log likelihood
+        void calculate_C();
+        void calculate_logl();
+
+        // Helper for perturb
+        double perturb_parameters(InfoNest::RNG& rng);
+
+    public:
+
+        // Generate from the distribution
+        void generate(InfoNest::RNG& rng);
+
+        // Override
+        void generate_data(InfoNest::RNG& rng);
+
+        // Printing to stream
+        void print(std::ostream& out) const;
+
+    public:
+        // A few options to use for `distance`
+        static double parameter_distance
+                            (const AR1& s1, const AR1& s2);
+
+};
+
+} // namespace WhittledAway
+
+#endif
+

### 2018-03-05 include/Model.h

@@ -35,7 +35,8 @@ class Model
         Eigen::LLT<Eigen::MatrixXd> L;
         Eigen::MatrixXd Lmat;
 
-        void generate_data(InfoNest::RNG& rng);
+        // In special cases, an override might help
+        virtual void generate_data(InfoNest::RNG& rng);
 
         // Member functions required
         virtual void calculate_C() = 0;

### 2018-03-05 src/AR1.cpp

@@ -1,0 +1,122 @@
+#include "AR1.h"
+#include "Utils.h"
+#include <cmath>
+#include <algorithm>
+#include <armadillo>
+#include <iostream>
+
+namespace WhittledAway
+{
+
+void AR1::generate(InfoNest::RNG& rng)
+{
+    mu = 1000.0*rng.randn();
+    beta = exp(-10.0 + 20.0*rng.rand());
+    L = exp(-10.0 + 20.0*rng.rand());
+
+    generate_data(rng);
+    calculate_logl();
+}
+
+void AR1::generate_data(InfoNest::RNG& rng)
+{
+    double alpha = exp(-1.0/L);
+    double sigma = beta/sqrt(1.0 - alpha*alpha);
+    y[0] = mu + sigma*rng.randn();
+    for(size_t i=1; i<N; ++i)
+        y[i] = mu + alpha*(y[i-1] - mu) + beta*rng.randn();
+}
+
+void AR1::calculate_C()
+{
+
+}
+
+double AR1::perturb_parameters(InfoNest::RNG& rng)
+{
+    double logH = 0.0;
+
+    int which = rng.rand_int(3);
+
+    if(which == 0)
+    {
+        logH -= -0.5*pow(mu/1000.0, 2);
+        mu += 1000.0*rng.randh();
+        logH += -0.5*pow(mu/1000.0, 2);
+    }
+    else if(which == 1)
+    {
+        beta = log(beta);
+        beta += 20.0*rng.randh();
+        InfoNest::wrap(beta, -10.0, 10.0);
+        beta = exp(beta);
+    }
+    else
+    {
+        L = log(L);
+        L += 20.0*rng.randh();
+        InfoNest::wrap(L, -10.0, 10.0);
+        L = exp(L);
+    }
+
+    calculate_logl();
+    return logH;
+}
+
+
+void AR1::calculate_logl()
+{
+    if(whittle)
+    {
+        logl = 0.0;
+
+        double model_psd, data_psd;
+
+        // Loop over first half of fft
+        for(size_t j=0; j<N/2; ++j)
+        {
+            // Model PSD
+//            model_psd = sigma*sigma;
+
+            // Data PSD
+            data_psd = (pow(y_fft[j].real(), 2) + pow(y_fft[j].imag(), 2));
+
+            // Whittle
+            logl += -log(model_psd) - data_psd/model_psd;
+        }
+    }
+    else
+    {
[…]

### 2018-03-05 src/main.cpp

@@ -4,13 +4,13 @@
 #include <type_traits>
 #include "Execute.hpp"
 #include "RNG.h"
-#include "Oscillation.h"
+#include "AR1.h"
 
 // MAIN
 int main()
 {
     // What example are we using? The class and the distance function.
-    using TheModel = WhittledAway::Oscillation;
+    using TheModel = WhittledAway::AR1;
 
     // Make sure TheModel is derived from WhittledAway::Model
     static_assert(std::is_base_of<WhittledAway::Model, TheModel>::value,

### 2018-03-05 src/AR1.cpp
Attempt at Whittle likelihood for AR(1) model

@@ -70,13 +70,17 @@ void AR1::calculate_logl()
     {
         logl = 0.0;
 
-        double model_psd, data_psd;
+        double model_psd, data_psd, f, w;
+        double alpha = exp(-1.0/L);
+        double sigma = beta/sqrt(1.0 - alpha*alpha);
 
         // Loop over first half of fft
         for(size_t j=0; j<N/2; ++j)
         {
+            f = static_cast<double>(j)/N;
+
             // Model PSD
-//            model_psd = sigma*sigma;
+            model_psd = sigma*sigma/(1.0 + alpha*alpha - 2*alpha*cos(2*M_PI*f));
 
             // Data PSD
             data_psd = (pow(y_fft[j].real(), 2) + pow(y_fft[j].imag(), 2));

### 2018-03-05 src/AR1.cpp
Minor edit

@@ -70,7 +70,7 @@ void AR1::calculate_logl()
     {
         logl = 0.0;
 
-        double model_psd, data_psd, f, w;
+        double model_psd, data_psd, f;
         double alpha = exp(-1.0/L);
         double sigma = beta/sqrt(1.0 - alpha*alpha);
 

### 2018-03-06 src/Oscillation.cpp
Use celerite for speedup

@@ -3,6 +3,7 @@
 #include <cmath>
 #include <algorithm>
 #include <armadillo>
+#include <celerite/celerite.h>
 #include <iostream>
 
 namespace WhittledAway
@@ -14,6 +15,7 @@ void Oscillation::generate(InfoNest::RNG& rng)
     log10_period = log10(N) - rng.rand();
     quality = exp(log(1.0) + log(1000.0)*rng.rand());
 
+    calculate_C();
     generate_data(rng);
     calculate_logl();
 }
@@ -73,8 +75,6 @@ double Oscillation::perturb_parameters(InfoNest::RNG& rng)
         quality = exp(quality);
     }
 
-    if(!whittle)
-        calculate_C();
     calculate_logl();
 
     return logH;
@@ -116,15 +116,49 @@ void Oscillation::calculate_logl()
         // Exact likelihood in the time domain
         logl = 0.0;
 
-        logl += -0.5*log(2*M_PI)*N;
+        // Only need these four
+        Eigen::VectorXd a(1);
+        Eigen::VectorXd b(1);
+        Eigen::VectorXd c(1);
+        Eigen::VectorXd d(1);
 
-        // Log determinant
-        double log_det = 0.0;
+        // Note: amplitude = S0*omega0*Q
+        double omega0, Q, Qterm;
+
+        omega0 = 2.0*M_PI/pow(10.0, log10_period);
+        Q = quality;
+
+        Qterm = sqrt(4*Q*Q - 1.0);
+        a(0) = A;
+        b(0) = A / Qterm;
+        c(0) = omega0 / (2*Q);
+        d(0) = c(0) * Qterm;
+
+        // Timestamps and variance
+        Eigen::VectorXd t(N), var(N);
         for(size_t i=0; i<N; ++i)
-            log_det += 2*log(Lmat(i, i));
+        {
+            t[i] = i;
+            var[i] = sigma*sigma;
+        }
 
-        // Solve
-        logl += -0.5*log_det - 0.5*y.dot(L.solve(y));
+        // Celerite solver
+        celerite::solver::CholeskySolver<double> solver;
+        try
+        {
+            solver.compute(0.0,
+                           Eigen::VectorXd(0), Eigen::VectorXd(0),
+                           a, b, c, d,
+                           t, var);
+
+            logl += -0.5*log(2*M_PI)*N;
+            logl += -0.5*solver.log_determinant();
+            logl += -0.5*solver.dot_solve(y);
+        }
[…]

### 2018-03-06 src/main.cpp

@@ -4,13 +4,13 @@
 #include <type_traits>
 #include "Execute.hpp"
 #include "RNG.h"
-#include "AR1.h"
+#include "Oscillation.h"
 
 // MAIN
 int main()
 {
     // What example are we using? The class and the distance function.
-    using TheModel = WhittledAway::AR1;
+    using TheModel = WhittledAway::Oscillation;
 
     // Make sure TheModel is derived from WhittledAway::Model
     static_assert(std::is_base_of<WhittledAway::Model, TheModel>::value,

### 2018-03-06 doc/ms.tex
New result

@@
 \hline
-20          &       Exact           & \\
+20          &       Exact           & $1.930 \pm 0.055$ \\
 20          &       Whittle         & $1.491 \pm 0.055$\\ 

### 2018-03-06 doc/ms.tex
One more result

--- 2018-03-05 doc/ms.tex
@@
 \hline
-20          &       Exact           & \\
-20          &       Whittle         & $1.491 \pm 0.055$\\ 
-50          &       Exact           & \\
+20          &       Exact           & $1.930 \pm 0.055$ \\
+50          &       Exact           & $2.391 \pm 0.056$ \\
+100         &       Exact           & $2.605 \pm 0.053$ \\
+200         &       Exact           &   \\
+20          &       Whittle         & $1.491 \pm 0.055$ \\ 
 50          &       Whittle         & $1.957 \pm 0.055$ \\
-100         &       Exact           & $2.605 \pm 0.053$ \\
 100         &       Whittle         & $2.052 \pm 0.055$ \\
-200         &       Exact           &   \\
 200         &       Whittle         &   \\

### 2018-03-06 doc/ms.tex
Abstract and intro writing

@@
 
-\abstract{\noindent I investigate the information content of a time series
-          dataset for inferring a parameter of interest. }
+\abstract{\noindent Inference from a time series dataset can be performed
+          either with the raw time series data, or after transformation to
+          the fourier domain. In the latter case, Whittle's approximation to
+          the likelihood function is computationally convenient. However, the
+          question of how much information is lost remains open --- in many
+          cases, the cost of extra computing time might be worth the benefits
+          of preserving all of the information from the raw time series data.
+          In this paper I investigate this question using numerical experiments
+          to quantify the mutual information between a dataset and a parameter
+          under the exact and the Whittle likelihoods.}
 
@@
 \section{Introduction}
+
+Just as probability theory (used in the `Bayesian' way)
+quantifies the degree to which one statement or
+proposition implies another \citep{knuth_skilling},
+information theory describes the
+degree to which resolving one question would resolve another
+\citep{knuth_questions, vanerp}.
 
@@
 
+\bibitem[Knuth(2005)]{knuth_questions}
+Knuth, K.H. Toward question-asking machines: The logic of questions and the inquiry calculus. In
+Proceedings of the 10th International Workshop On Artificial Intelligence and Statistics, Barbados, 6–8
+January 2005.
+
+\bibitem[Knuth and Skilling(2012)]{knuth_skilling}
+Knuth, Kevin H., and John Skilling. ``Foundations of inference.''
+Axioms 1, no. 1 (2012): 38-73.
+
 \bibitem[Lomb(1976)]{lomb}
 Lomb, Nicholas R. ``Least-squares frequency analysis of unequally spaced data.'' Astrophysics and space science 39, no. 2 (1976): 447-462.
+
+\bibitem[van Erp et al.(2017)]{vanerp}
+van Erp, H. R., Ronald O. Linger, and Pieter HAJM van Gelder. ``Inquiry Calculus and the Issue of Negative Higher Order Informations.'' Entropy 19, no. 11 (2017): 622.
 

### 2018-03-06 doc/ms.tex
Discussion of priors, and more info in tables

@@
 \end{align}
+The prior for the amplitude was chosen so that it is likely to be of the same
+order of magnitude as the noise level, and the prior for the period asserts that
+there are between one and ten complete periods within the observation period.
+I did not allow shorter periods than this, so the current paper does not
+address the issue of aliasing (equivalently, multimodal posterior distributions
+for the period).
+
 We proceeded to compute the conditional entropy
@@
 \hline
-50          & \\
+20          &       $0.439 \pm 0.053$ \\
+50          &       $0.434 \pm 0.056$
 100         &       $0.553 \pm 0.056$ \\

### 2018-03-06 doc/ms.tex
Fixed error
Wrote down the whittle likelihood

@@
 usually only accessible through sampling, such as posterior distributions.
+
+\subsection{Time series and the whittle likelihood}
+
+Assume a quantity has been measured at $N$ equally-spaced time points
+$\boldsymbol{t} = \{0, 1, 2, ..., N-1\}$, resulting in observations
+$\boldsymbol{y} = \{y_0, y_1, ..., y_{N-1}\}$.
+% Use GP framework
+
+The whittle likelihood is
+\begin{align}
+p(\boldsymbol{y} | \boldsymbol{\theta})
+    &\propto -\sum_{k=0}^{N/2 - 1}
+         \left(S(\omega_k; \boldsymbol{\theta})
+         + \frac{I(\omega_k; \boldsymbol{y})}
+                {S(\omega_k; \boldsymbol{\theta})}\right)\label{eqn:whittle}
+\end{align}
+where $I(\omega_k; \boldsymbol{y})$ is the power spectrum of the data,
+$S(\omega_k; \boldsymbol{\theta})$ is the power spectral density implied
+by the model parameters, and the summation is over the angular frequencies
+involved in the first half of the discrete fourier transform.
+Equation~\ref{eqn:whittle} is equivalent to an exponential distribution
+for the empirical power spectrum
+$I(\omega_k; \boldsymbol{y})$ whose expectation is given by the
+model power spectrum $S(\omega_k; \boldsymbol{\theta})$.
 
@@
 20          &       $0.439 \pm 0.053$ \\
-50          &       $0.434 \pm 0.056$
+50          &       $0.434 \pm 0.056$ \\
 100         &       $0.553 \pm 0.056$ \\

### 2018-03-06 doc/ms.tex
New results and writing

@@
 \usepackage{graphicx}
+\usepackage{hyperref}
 \usepackage[utf8]{inputenc}
@@
 $\boldsymbol{y} = \{y_0, y_1, ..., y_{N-1}\}$.
-% Use GP framework
+Throughout this paper, I will specify all models in the framework of
+Gaussian processes. Ultimately, this means the probability distribution for
+the data $\boldsymbol{y}$ given some parameters $\boldsymbol{\theta}$ will
+be a multivariate normal distribution whose mean vector $\boldsymbol{\mu}$
+and covariance matrix $\mathbf{C}$ depend on $\boldsymbol{\theta}$:
+\begin{align}
+\boldsymbol{y} | \boldsymbol{\theta} &\sim
+    \textnormal{Normal}\left(\boldsymbol{\mu}(\boldsymbol{\theta}),
+                             \mathbf{C}(\boldsymbol{\theta})\right).
+\end{align}
+The full expression for this probability density is
+\begin{align}
+p(\boldsymbol{y} | \boldsymbol{\theta})
+    &= \frac{1}{\sqrt{(2\pi)^N\det \mathbf{C}}}
+        \exp\left[-\frac{1}{2}
+                    (\boldsymbol{y} - \boldsymbol{\mu})^T
+                    \mathbf{C}^{-1}
+                    (\boldsymbol{y} - \boldsymbol{\mu})\right]
+\end{align}
+which also provides the exact likelihood function if we wish to infer
+$\boldsymbol{\theta}$ from $\boldsymbol{y}$. The determinant and inverse
+are typically expensive, though in some cases fast methods are available
+\citep[e.g.][]{mackay, hodlr, celerite}.
 
@@
     &\propto -\sum_{k=0}^{N/2 - 1}
-         \left(S(\omega_k; \boldsymbol{\theta})
+         \left[S(\omega_k; \boldsymbol{\theta})
          + \frac{I(\omega_k; \boldsymbol{y})}
-                {S(\omega_k; \boldsymbol{\theta})}\right)\label{eqn:whittle}
+                {S(\omega_k; \boldsymbol{\theta})}\right]\label{eqn:whittle}
 \end{align}
@@
 
-Methods that analyse the time series data directly are often less convenient than frequency-domain fits, and can have a much greater computational cost,
+Methods that analyse the time series data directly are often less convenient
+than frequency-domain fits, and can have a much greater computational cost,
 despite recent computational advances which have improved the matter
@@
 100         &       Exact           & $2.605 \pm 0.053$ \\
-200         &       Exact           &   \\
+200         &       Exact           & $2.830 \pm 0.054$ \\
 20          &       Whittle         & $1.491 \pm 0.055$ \\ 
@@
 100         &       Whittle         & $2.052 \pm 0.055$ \\
-200         &       Whittle         &   \\
+200         &       Whittle         & $2.158 \pm 0.061$  \\
 \bottomrule
@@
 100         &       $0.553 \pm 0.056$ \\
-200         & \\
+200         &       $0.671 \pm 0.062$ \\
 \bottomrule
@@
 
+\bibitem[Gibbs and MacKay(1997)]{mackay}
+MacKay, D., and M. N. Gibbs. ``Efficient implementation of Gaussian processes.'' Neural Computation (1997).
+Available at \url{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.48.8224&rep=rep1&type=pdf}
+(as of 2018-03-07).
+
 \bibitem[van Erp et al.(2017)]{vanerp}

### 2018-03-07 src/Oscillation.cpp
Don't print data

@@ -168,8 +168,8 @@ void Oscillation::calculate_logl()
 void Oscillation::print(std::ostream& out) const
 {
     out<<A<<' '<<log10_period<<' '<<quality<<' ';
-    for(size_t i=0; i<N; ++i)
-        out << y[i] << ' ';
+//    for(size_t i=0; i<N; ++i)
+//        out << y[i] << ' ';
 }
 
 double Oscillation::parameter_distance(const Oscillation& s1, const Oscillation& s2)

### 2018-03-07 test/test.cpp

@@ -1,0 +1,50 @@
+// Includes
+#include <iostream>
+#include "Oscillation.h"
+#include "RNG.h"
+
+using namespace WhittledAway;
+
+// MAIN
+int main()
+{
+    // What example are we using? The class and the distance function.
+    using TheModel = Oscillation;
+
+    // Make sure TheModel is derived from WhittledAway::Model
+    static_assert(std::is_base_of<Model, TheModel>::value,
+                    "Examples must derive from WhittledAway::Model.");
+
+    // Create random number generator
+    InfoNest::RNG rng(0);
+
+    // Make a particle
+    TheModel o1;
+    o1.generate(rng);
+
+    // Print the particle
+    o1.print(std::cout);
+    std::cout << std::endl;
+
+    for(int i=0; i<1000000; ++i)
+    {
+        // Make a proposal
+        auto o2 = o1;
+        double logH = o2.perturb(rng);
+
+        // Accept?
+        if(rng.rand() <= exp(logH))
+            o1 = o2;
+
+        // Print
+        if((i+1)%100 == 0)
+        {
+            std::cerr << (i+1) << std::endl;
+            o1.print(std::cout);
+            std::cout << std::endl;
+        }
+    }
+
+    return 0;
+}
+

### 2018-03-07 src/Model.cpp
Heaps of work

@@ -35,7 +35,7 @@ void Model::generate_data(InfoNest::RNG& rng)
         y_fft[i] = y[i];
 
     // Take the fft
-    y_fft = arma::fft(y_fft)/sqrt(N);
+    y_fft = arma::fft(y_fft);
 }
 
 

### 2018-03-07 src/Oscillation.cpp

@@ -25,7 +25,6 @@ void Oscillation::calculate_C()
     // Fill covariance matrix
     double w0 = 2*M_PI/pow(10.0, log10_period);
     double eta = sqrt(std::abs(1.0 - 1.0/(4.0*quality*quality)));
-    double S0 = A/w0/quality;
     double tau;
 
     for(size_t i=0; i<N; ++i)
@@ -34,7 +33,7 @@ void Oscillation::calculate_C()
         {
             tau = std::abs(static_cast<double>(i) - static_cast<double>(j));
             C(i, j) = cos(eta*w0*tau) + sin(eta*w0*tau)/(2.0*eta*quality);
-            C(i, j) *= S0*w0*quality*exp(-w0*tau/(2.0*quality));
+            C(i, j) *= A*A*exp(-w0*tau/(2.0*quality));
 
             if(i != j)
                 C(j, i) = C(i, j);
@@ -54,6 +53,7 @@ double Oscillation::perturb_parameters(InfoNest::RNG& rng)
 
     // Perturb one of the three parameters
     int which = rng.rand_int(3);
+
     if(which == 0)
     {
         A = log(A);
@@ -85,30 +85,28 @@ void Oscillation::calculate_logl()
 {
     if(whittle)
     {
-
-        // https://www4.stat.ncsu.edu/~reich/SpatialStats/code/Whittle.html
         logl = 0.0;
 
-        double model_psd, data_psd, f, w;
+        double model_psd, data_pgram, f, w;
         double w0 = 2*M_PI/pow(10.0, log10_period);
-        double S0 = A/w0/quality;
-        double coeff = sqrt(2.0/M_PI)*S0*pow(w0, 4);
+        double S0 = A*A/w0/quality;
+        double coeff = 2.0*S0*pow(w0, 4);
 
-        // Loop over first half of fft
-        for(size_t j=0; j<N/2; ++j)
+        // Loop over positive frequencies
+        for(size_t j=1; j<=(size_t)floor(0.5*(N-1)); ++j)
         {
             // Model PSD (Celerite paper, Eqn 20)
-            f = static_cast<double>(j) / N;
+            f = static_cast<double>(j)/N;
             w = 2*M_PI*f;
             model_psd = coeff/(pow(w*w - w0*w0, 2)
                                         + w0*w0*w*w/(quality*quality));
             model_psd += sigma*sigma;
 
-            // Data PSD
[…]

### 2018-03-07 src/WhiteNoise.cpp

@@ -62,19 +62,19 @@ void WhiteNoise::calculate_logl()
     {
         logl = 0.0;
 
-        double model_psd, data_psd;
+        double model_psd, data_pgram;
 
         // Loop over first half of fft
-        for(size_t j=0; j<N/2; ++j)
+        for(size_t j=1; j<=(size_t)floor(0.5*(N-1)); ++j)
         {
             // Model PSD
             model_psd = sigma*sigma;
 
-            // Data PSD
-            data_psd = (pow(y_fft[j].real(), 2) + pow(y_fft[j].imag(), 2));
+            // Data periodogram
+            data_pgram = (pow(y_fft[j].real(), 2) + pow(y_fft[j].imag(), 2)) / N;
 
             // Whittle
-            logl += -log(model_psd) - data_psd/model_psd;
+            logl += -log(model_psd) - data_pgram/model_psd;
         }
     }
     else

### 2018-03-07 test/test.cpp

@@ -16,7 +16,7 @@ int main()
                     "Examples must derive from WhittledAway::Model.");
 
     // Create random number generator
-    InfoNest::RNG rng(0);
+    InfoNest::RNG rng(1);
 
     // Make a particle
     TheModel o1;
@@ -26,7 +26,8 @@ int main()
     o1.print(std::cout);
     std::cout << std::endl;
 
-    for(int i=0; i<1000000; ++i)
+    // Do MCMC
+    for(int i=0; i<10000000; ++i)
     {
         // Make a proposal
         auto o2 = o1;
@@ -36,14 +37,14 @@ int main()
         if(rng.rand() <= exp(logH))
             o1 = o2;
 
-        // Print
         if((i+1)%100 == 0)
         {
             std::cerr << (i+1) << std::endl;
             o1.print(std::cout);
             std::cout << std::endl;
         }
-    }
+}
+
 
     return 0;
 }

### 2018-03-07 doc/ms.tex
Fixed errors; latest thinking on S(omega)
fixing up whittle formula

@@
 p(\boldsymbol{y} | \boldsymbol{\theta})
-    &\propto -\sum_{k=0}^{N/2 - 1}
+    &\propto \exp\left(-\sum_{k=1}^{\lfloor N/2 - 1 \rfloor}
          \left[S(\omega_k; \boldsymbol{\theta})
          + \frac{I(\omega_k; \boldsymbol{y})}
-                {S(\omega_k; \boldsymbol{\theta})}\right]\label{eqn:whittle}
+                {S(\omega_k; \boldsymbol{\theta})}\right]
+                \right)
+\label{eqn:whittle}
 \end{align}
@@
 Equation~\ref{eqn:whittle} is equivalent to an exponential distribution
-for the empirical power spectrum
+for the empirical power spectrum (i.e., the periodogram)
 $I(\omega_k; \boldsymbol{y})$ whose expectation is given by the
-model power spectrum $S(\omega_k; \boldsymbol{\theta})$.
+model power spectrum (i.e., the power spectral density)
+$S(\omega_k; \boldsymbol{\theta})$.
 
@@
 \frac{d^2y}{dt^2} + \frac{\omega_0}{Q}\frac{dy}{dt}
-        + \omega_0^2\frac{dy}{dt} &= \epsilon(t).\label{eqn:ode}
+        + \omega_0^2y &= \epsilon(t).\label{eqn:ode}
 \end{align}
@@
 \begin{align}
-C(\tau) &= A^2\left[\cos(\eta\omega_0\tau)
+C(\tau) &= A^2\exp\left(-\frac{\omega_0\tau}{2Q}\right)
+            \left[\cos(\eta\omega_0\tau)
                 + \frac{1}{2\eta Q}\sin(\eta\omega_0\tau)\right]
@@
 \begin{align}
-S(\omega) &= \sqrt{\frac{2}{\pi}}
-             \frac{S_0\omega_0^4}
+S(\omega) &= \frac{2S_0\omega_0^4}
                   {(\omega^2 - \omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
 \end{align}
-where $S_0 = A/(\omega_0Q)$.
+where $S_0 = A^2/(\omega_0Q)$. {\bf NB: different from Celerite paper
+due to changed Fourier Transform convention}
 

### 2018-03-07 doc/ms.tex
New results after fixing bugs (needs more testing though)
Explain the factor of sqrt(2pi)

@@
 \end{align}
-where $S_0 = A^2/(\omega_0Q)$. {\bf NB: different from Celerite paper
-due to changed Fourier Transform convention}
+where $S_0 = A^2/(\omega_0Q)$. This differs from the formula given by
+\citet{celerite} by a factor of $\sqrt{2\pi}$ due to different fourier
+transform conventions (this formula omits a factor of $\frac{1}{\sqrt{2\pi}$
+in front of the continuous fourier transform).
 
@@
 \hline
-20          &       Exact           & $1.930 \pm 0.055$ \\
-50          &       Exact           & $2.391 \pm 0.056$ \\
-100         &       Exact           & $2.605 \pm 0.053$ \\
-200         &       Exact           & $2.830 \pm 0.054$ \\
-20          &       Whittle         & $1.491 \pm 0.055$ \\ 
-50          &       Whittle         & $1.957 \pm 0.055$ \\
-100         &       Whittle         & $2.052 \pm 0.055$ \\
-200         &       Whittle         & $2.158 \pm 0.061$  \\
+100         &       Exact           & $2.488 \pm 0.056$ \\
+100         &       Whittle         & $2.007 \pm 0.056$ \\
 \bottomrule
@@
 \hline
-20          &       $0.439 \pm 0.053$ \\
-50          &       $0.434 \pm 0.056$ \\
-100         &       $0.553 \pm 0.056$ \\
-200         &       $0.671 \pm 0.062$ \\
+100         &       $0.481 \pm 0.057$ \\
 \bottomrule

### 2018-03-07 doc/ms.tex
Tidied up mistake

@@
 \citet{celerite} by a factor of $\sqrt{2\pi}$ due to different fourier
-transform conventions (this formula omits a factor of $\frac{1}{\sqrt{2\pi}$
+transform conventions (this formula omits a factor of $1/\sqrt{2\pi}$
 in front of the continuous fourier transform).

### 2018-03-07 doc/ms.tex
Mean vector to zero

@@
 be a multivariate normal distribution whose mean vector $\boldsymbol{\mu}$
-and covariance matrix $\mathbf{C}$ depend on $\boldsymbol{\theta}$:
+and covariance matrix $\mathbf{C}$ depend on $\boldsymbol{\theta}$. Throughout
+this paper the mean vector will be set to zero, so dependence on the parameters
+is through the covariance matrix only:
 \begin{align}
 \boldsymbol{y} | \boldsymbol{\theta} &\sim
-    \textnormal{Normal}\left(\boldsymbol{\mu}(\boldsymbol{\theta}),
+    \textnormal{Normal}\left(\boldsymbol{0},
                              \mathbf{C}(\boldsymbol{\theta})\right).
@@
         \exp\left[-\frac{1}{2}
-                    (\boldsymbol{y} - \boldsymbol{\mu})^T
+                    \boldsymbol{y}^T
                     \mathbf{C}^{-1}
-                    (\boldsymbol{y} - \boldsymbol{\mu})\right]
+                    \boldsymbol{y}\right]
 \end{align}

### 2018-03-07 doc/ms.tex
Nonzero frequencies

@@
 $S(\omega_k; \boldsymbol{\theta})$ is the power spectral density implied
-by the model parameters, and the summation is over the angular frequencies
+by the model parameters, and the summation is over the nonzero angular frequencies
 involved in the first half of the discrete fourier transform.

### 2018-03-07 doc/ms.tex
Cite Matt and Renate

@@
 directly from the time series data, without any use of the frequency domain
-\citep[e.g.][]{gregory_loredo, gregory, kelly, brewer_stello, farr}.
+\citep[e.g.][]{gregory_loredo, gregory, kelly, brewer_stello, farr, edwards}.
 
@@
 
+\bibitem[Edwards et al.(2015)]{edwards}
+Edwards, Matthew C., Renate Meyer, and Nelson Christensen.
+``Bayesian semiparametric power spectral density estimation with applications
+in gravitational wave data analysis.'' Physical Review D 92, no. 6 (2015): 064011.
+
 \bibitem[Farr et al.(2018)]{farr}

### 2018-03-07 src/Oscillation.cpp
Older solution put in inside a comment

@@ -75,6 +75,7 @@ double Oscillation::perturb_parameters(InfoNest::RNG& rng)
         quality = exp(quality);
     }
 
+//    calculate_C();
     calculate_logl();
 
     return logH;
@@ -155,6 +156,21 @@ void Oscillation::calculate_logl()
         {
             logl = -1E300;
         }
+
+        /*
+        // Non-celerite computation, tested and found to be equivalent
+        logl = 0.0;
+        logl += -0.5*log(2*M_PI)*N;
+
+        // Log determinant
+        double log_det = 0.0;
+        for(size_t i=0; i<N; ++i)
+            log_det += 2*log(Lmat(i, i));
+
+        // Solve
+        logl += -0.5*log_det - 0.5*y.dot(L.solve(y));
+        */
+
     }
 
     if(std::isnan(logl) || std::isinf(logl))

### 2018-03-07 include/WhiteNoise.h
Speed up White Noise

@@ -19,7 +19,7 @@ class WhiteNoise : public Model
         // Noise sd
         double sigma;
 
-        // Calculate log likelihood
+        void generate_data(InfoNest::RNG& rng);
         void calculate_C();
         void calculate_logl();
 

### 2018-03-07 src/WhiteNoise.cpp

@@ -12,9 +12,14 @@ void WhiteNoise::generate(InfoNest::RNG& rng)
 {
     sigma = exp(-10.0 + 20.0*rng.rand());
 
-    calculate_C();
     generate_data(rng);
     calculate_logl();
+}
+
+void WhiteNoise::generate_data(InfoNest::RNG& rng)
+{
+    for(size_t i=0; i<N; ++i)
+        y[i] = sigma*rng.randn();
 }
 
 void WhiteNoise::calculate_C()

### 2018-03-08 src/WhiteNoise.cpp
Custom generate_data needs to take the fft

@@ -20,6 +20,13 @@ void WhiteNoise::generate_data(InfoNest::RNG& rng)
 {
     for(size_t i=0; i<N; ++i)
         y[i] = sigma*rng.randn();
+
+    // Copy into the Armadillo vector
+    for(size_t i=0; i<N; ++i)
+        y_fft[i] = y[i];
+
+    // Take the fft
+    y_fft = arma::fft(y_fft);
 }
 
 void WhiteNoise::calculate_C()

### 2018-03-09 include/AR1.h
Got sympy to calculate the PSD for AR(1). Untested though

@@ -16,8 +16,8 @@ class AR1 : public Model
 {
     private:
 
-        // Mean level, beta, L
-        double mu, beta, L;
+        // Innovation SD and timescale
+        double beta, L;
 
         // Calculate log likelihood
         void calculate_C();

### 2018-03-09 src/AR1.cpp

@@ -10,7 +10,6 @@ namespace WhittledAway
 
 void AR1::generate(InfoNest::RNG& rng)
 {
-    mu = 1000.0*rng.randn();
     beta = exp(-10.0 + 20.0*rng.rand());
     L = exp(-10.0 + 20.0*rng.rand());
 
@@ -22,9 +21,16 @@ void AR1::generate_data(InfoNest::RNG& rng)
 {
     double alpha = exp(-1.0/L);
     double sigma = beta/sqrt(1.0 - alpha*alpha);
-    y[0] = mu + sigma*rng.randn();
+    y[0] = sigma*rng.randn();
     for(size_t i=1; i<N; ++i)
-        y[i] = mu + alpha*(y[i-1] - mu) + beta*rng.randn();
+        y[i] = alpha*y[i-1] + beta*rng.randn();
+
+    // Copy into the Armadillo vector
+    for(size_t i=0; i<N; ++i)
+        y_fft[i] = y[i];
+
+    // Take the fft
+    y_fft = arma::fft(y_fft);
 }
 
 void AR1::calculate_C()
@@ -36,15 +42,9 @@ double AR1::perturb_parameters(InfoNest::RNG& rng)
 {
     double logH = 0.0;
 
-    int which = rng.rand_int(3);
+    int which = rng.rand_int(2);
 
     if(which == 0)
-    {
-        logH -= -0.5*pow(mu/1000.0, 2);
-        mu += 1000.0*rng.randh();
-        logH += -0.5*pow(mu/1000.0, 2);
-    }
-    else if(which == 1)
     {
         beta = log(beta);
         beta += 20.0*rng.randh();
@@ -70,23 +70,26 @@ void AR1::calculate_logl()
     {
         logl = 0.0;
 
-        double model_psd, data_psd, f;
+        double model_psd, data_pgram, w;
         double alpha = exp(-1.0/L);
         double sigma = beta/sqrt(1.0 - alpha*alpha);
 
         // Loop over first half of fft
-        for(size_t j=0; j<N/2; ++j)
+        for(size_t j=1; j<=(size_t)floor(0.5*(N-1)); ++j)
         {
-            f = static_cast<double>(j)/N;
+            // Angular frequency
+            w = 2.0*M_PI*static_cast<double>(j)/N;
 
             // Model PSD
-            model_psd = sigma*sigma/(1.0 + alpha*alpha - 2*alpha*cos(2*M_PI*f));
+            model_psd = 2.0*sigma*sigma*L
[…]

### 2018-03-09 doc/psd_sandpit.py
PSD script

@@ -1,0 +1,41 @@
+from numba import jit
+from pylab import *
+
+w0 = 0.6
+Q = 2.0 #5.0
+eta = sqrt(1 - 1/(4*Q**2))
+A = 2.1 #3.5
+L = 30.0
+
+t = linspace(-300, 300, 10001)
+C = A**2*exp(-w0*abs(t)/(2*Q))*(cos(eta*w0*abs(t)) + sin(eta*w0*abs(t))/(2*eta*Q))
+C = A**2*exp(-abs(t)/L)
+
+@jit
+def fourier_transform(f, w):
+    F = zeros(len(w))
+    for i in range(len(w)):
+        F[i] = real(trapz(f*exp(-1j*w[i]*t), x=t))
+        print(i+1)
+    return F
+
+figure(1)
+plot(t, C)
+xlabel("$t$")
+ylabel("Covariance")
+
+w = linspace(-10, 10, 10001)
+S0 = A**2/(w0*Q)
+#S = 2*S0*w0**4/((w**2 - w0**2)**2 + w0**2*w**2/Q**2)
+S = 2.0*A*A*L/(L*L*w*w + 1.0);
+
+F = fourier_transform(C, w)
+print("Ratio = {r}".format(r=(F/S)[len(F)//2]))
+
+figure(2)
+plot(w, F)
+plot(w, S)
+xlabel("$\\omega$")
+ylabel("PSD")
+show()
+

### 2018-03-09 src/AR1.cpp
Fixed formula?

@@ -82,7 +82,7 @@ void AR1::calculate_logl()
 
             // Model PSD
             model_psd = 2.0*sigma*sigma*L
-                            /(4.0*M_PI*M_PI*L*L*w*w + 1.0);
+                            /(L*L*w*w + 1.0);
 
             // Data periodogram
             data_pgram = (pow(y_fft[j].real(), 2)

### 2018-03-10 src/AR1.cpp
Changed prior on L

@@ -11,7 +11,7 @@ namespace WhittledAway
 void AR1::generate(InfoNest::RNG& rng)
 {
     beta = exp(-10.0 + 20.0*rng.rand());
-    L = exp(-10.0 + 20.0*rng.rand());
+    L = exp(log(0.1) + log(1E5)*rng.rand());
 
     generate_data(rng);
     calculate_logl();
@@ -54,8 +54,8 @@ double AR1::perturb_parameters(InfoNest::RNG& rng)
     else
     {
         L = log(L);
-        L += 20.0*rng.randh();
-        InfoNest::wrap(L, -10.0, 10.0);
+        L += log(1E5)*rng.randh();
+        InfoNest::wrap(L, log(0.1), log(1E4));
         L = exp(L);
     }
 

### 2018-03-10 src/AR1.cpp
Distance function changed

@@ -122,7 +122,10 @@ void AR1::print(std::ostream& out) const
 
 double AR1::parameter_distance(const AR1& s1, const AR1& s2)
 {
-    return std::abs(log(s2.L) - log(s1.L));
+    double dsq = 0.0;
+    dsq += pow(log(s1.beta) - log(s2.beta), 2);
+    dsq += pow(log(s1.L) - log(s2.L), 2);
+    return sqrt(dsq);
 }
 
 } // namespace WhittledAway

### 2018-03-10 src/AR1.cpp
Don't allow such long timescales

@@ -11,7 +11,7 @@ namespace WhittledAway
 void AR1::generate(InfoNest::RNG& rng)
 {
     beta = exp(-10.0 + 20.0*rng.rand());
-    L = exp(log(0.1) + log(1E5)*rng.rand());
+    L = exp(log(0.1) + log(10.0*N)*rng.rand());
 
     generate_data(rng);
     calculate_logl();
@@ -54,8 +54,8 @@ double AR1::perturb_parameters(InfoNest::RNG& rng)
     else
     {
         L = log(L);
-        L += log(1E5)*rng.randh();
-        InfoNest::wrap(L, log(0.1), log(1E4));
+        L += log(10.0*N)*rng.randh();
+        InfoNest::wrap(L, log(0.1), log(N));
         L = exp(L);
     }
 

### 2018-03-13 src/AR1.cpp
Correct AR(1) formula

@@ -72,7 +72,9 @@ void AR1::calculate_logl()
 
         double model_psd, data_pgram, w;
         double alpha = exp(-1.0/L);
+        double alpha_sq = alpha*alpha;
         double sigma = beta/sqrt(1.0 - alpha*alpha);
+        double sigma_sq = sigma*sigma;
 
         // Loop over first half of fft
         for(size_t j=1; j<=(size_t)floor(0.5*(N-1)); ++j)
@@ -81,8 +83,8 @@ void AR1::calculate_logl()
             w = 2.0*M_PI*static_cast<double>(j)/N;
 
             // Model PSD
-            model_psd = 2.0*sigma*sigma*L
-                            /(L*L*w*w + 1.0);
+            model_psd = sigma_sq*(1.0 - alpha_sq)
+                            / (1.0 + alpha_sq - 2.0*alpha*cos(w)); 
 
             // Data periodogram
             data_pgram = (pow(y_fft[j].real(), 2)

### 2018-03-13 test/test.cpp

@@ -1,4 +1,5 @@
 // Includes
+#include <ctime>
 #include <iostream>
 #include "Models.h"
 #include "RNG.h"
@@ -9,14 +10,14 @@ using namespace WhittledAway;
 int main()
 {
     // What example are we using? The class and the distance function.
-    using TheModel = Oscillation;
+    using TheModel = AR1;
 
     // Make sure TheModel is derived from WhittledAway::Model
     static_assert(std::is_base_of<Model, TheModel>::value,
                     "Examples must derive from WhittledAway::Model.");
 
     // Create random number generator
-    InfoNest::RNG rng(1);
+    InfoNest::RNG rng(time(0));
 
     // Make a particle
     TheModel o1;

### 2018-03-14 doc/ms.tex
A^2
More results
PSD of AR(1)
AR(1) writing

@@
 
+\section{Example 1: An AR(1) process}
+Consider a continuous AR(1) process, whose covariance function is
+\begin{align}
+C(\tau) &= \sigma^2 \exp\left(-\frac{|\tau|}{L}\right).
+\end{align}
+If the process is observed at times $\boldsymbol{t} = \{t_0, ..., t_{N-1}\}$,
+the distribution of the discrete distribution for the $\boldsymbol{y}$
+is a discrete AR(1) distribution with covariance of the same form.
+The joint distribution for $\boldsymbol{y}$ given the parameters
+may also be written
+\begin{align}
+y_i &= \alpha y_{i-1} + \beta n_i
+\end{align}
+where $\alpha = \exp(-1/L)$, $\beta=\sigma\sqrt{1-\alpha^2}$
+is the standard deviation of the
+innovations, and the $\{n_i\}$ quantities have independent Normal$(0, 1)$
+distributions.
+
+The power spectral density (PSD) of the discrete distribution is
+\begin{align}
+S(\omega) &= \sigma^2\frac{1 - \alpha^2}{1 + \alpha^2 - 2\alpha\cos(\omega)}.
+\end{align}
+
 \section{A Damped, Stochastically-Excited Oscillator}
@@
 \begin{align}
-C(\tau) &= A\exp\left(-\frac{\omega_0\tau}{2Q}\right)\cos(\omega_0\tau)
+C(\tau) &= A^2\exp\left(-\frac{\omega_0\tau}{2Q}\right)\cos(\omega_0\tau)
 \end{align}
@@
 \hline
+20          &       Exact           & $1.945 \pm 0.057$ \\
+50          &       Exact           & $2.263 \pm 0.058$ \\
 100         &       Exact           & $2.488 \pm 0.056$ \\
+200         &       Exact           & $2.750 \pm 0.054$ \\
+\hline
+20          &       Whittle         & $1.552 \pm 0.048$ \\
+50          &       Whittle         & $1.863 \pm 0.056$ \\
 100         &       Whittle         & $2.007 \pm 0.056$ \\
+200         &       Whittle         & $2.089 \pm 0.060$ \\
 \bottomrule
@@
 \hline
+20          &       $0.393 \pm 0.050$ \\
+50          &       $0.400 \pm 0.056$ \\
 100         &       $0.481 \pm 0.057$ \\
+200         &       $0.661 \pm 0.061$ \\
 \bottomrule
@@
 
+\appendix
+\section{Notes}
+
+The discrete AR(1) process has covariance function
+\begin{align}
+C(\tau) &= \sigma^2 \exp\left(-|\tau| / L\right) \\
+        &= \sigma^2 \alpha^{|\tau|},
+\end{align}
+where $\tau$ is the {\em integer} lag.
+The power spectral density is therefore
+\begin{align}
+S(\omega) &= \sum_{\tau=-\infty}^{\infty} C(\tau)\exp(-i \omega \tau) \\
+          &= \sum_{\tau=-\infty}^{\infty} \sigma^2 \alpha^{|\tau|}
+                                         \exp(-i \omega \tau) \\
+          &= \sigma^2 \left(
+                \sum_{\tau=-\infty}^{-1} \alpha^{-\tau}\exp(-i \omega \tau) + 
+                \sum_{\tau=0}^{\infty} \alpha^{\tau}\exp(-i \omega \tau)
+                        \right) \\
+          &= \sigma^2 \left(
+                \frac{\alpha e^{i\omega}}
+                     {1 - \alpha e^{i\omega}}
+                 +
+                \frac{1}
+                     {1 - \alpha e^{-i\omega}}
+                        \right) \\
+         &= \sigma^2
+                \frac{1 - \alpha^2}
+                     {1 + \alpha^2 - 2\alpha\cos(\omega)}
+\end{align}
+
 \end{document}

### 2018-03-21 doc/ms.tex

@@
 $S(\omega_k; \boldsymbol{\theta})$.
+
+If the {\em discrete} covariance function, defined for {\em integer} lag
+$\tau$, is $C(\tau)$
+then the PSD is a {\em continuous} function of angular frequency
+$\omega \in (-\pi, \pi]$,
+defined as
+\begin{align}
+S(\omega) &= \sum_\tau C(\tau)\exp\left(-i \omega \tau\right).
+\end{align}
+Since $C$ is necessarily an even function of $\tau$, the imaginary part of
+$S(\omega)$ will be zero. In the Whittle likelihood, $S(\omega)$ is then
+evaluated at a discrete set of points.
 
